Machine Learning Driven Set of microRNAs as a Novel Biomarker Set for Myocardial Infarction Diagnosis

### 2023-08-07 ArticleDraft.Rmd
The version after revision

---
title: Machine Learning Driven Set of microRNAs as a Novel Biomarker Set for Myocardial Infarction Diagnosis
authors:
  - name: "Mehrdad Samadishadlou"
    address: "Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran"
date: "`r Sys.Date()`"
link-citation: yes
bibliography: [For Article Refrence.bib, discussion.bib, Introduction.bib, mir296.bib]
abstract: |
  MicroRNAs (miRNAs) play a crucial role in regulating adaptive and maladaptive responses in cardiovascular diseases, making them attractive targets for potential biomarkers. However, their potential as novel biomarkers for diagnosing cardiovascular diseases requires systematic evaluation. In this study, we aimed to identify a key set of miRNA biomarkers using integrated bioinformatics and machine learning analysis. We combined and analyzed three gene expression datasets from the Gene Expression Omnibus (GEO) database, which contains peripheral blood mononuclear cells (PBMCs) samples from individuals with myocardial infarction (MI), stable coronary artery disease (CAD), and healthy individuals. Additionally, we selected a set of miRNAs based on their area under the receiver operating characteristic curve (AUC-ROC) for separating the CAD and MI samples. We designed a two-layer architecture for sample classification, in which the first layer isolates healthy samples from not-healthy ones, and the second layer classifies stable CAD and MI samples. We trained different machine learning models using both biomarker sets and evaluated their performance on a test set. We identified miR-21, miR-186, and miR-32 as the only miRNAs among the differentially expressed genes, and a set including miR-186, miR-21, miR-197, miR-29A, and miR-296 as the optimum set of miRNAs selected by their AUC-ROC. Both biomarker sets could distinguish healthy from not-healthy samples with complete accuracy. The best performance for the classification of CAD and MI was achieved with an SVM model trained using the biomarker set selected by AUC-ROC, with an AUC-ROC of 0.96 and an accuracy of 0.94 on the test data. Our study demonstrated that miRNA signatures derived from PBMCs could serve as valuable novel biomarkers for cardiovascular diseases.

output: 
  bookdown::pdf_book:
  bookdown::word_document2: default
always_allow_html: true
---

@@

At present, cardiovascular diseases (CVDs) are the leading cause of human mortality, with 32% of all global deaths. It is estimated that about 85% of CVDs mortality were diagnosed with myocardial infarction (MI) [@CVD]. MI is an acute coronary syndrome with sudden blockage and stenosis of the coronary artery and subsequent myocardial ischemia, leading to extensive cardiomyocyte damage and necrosis [@Macro].

Over the last 50 years, numerous attempts have been collected to use biomarkers to facilitate diagnosis, assess the risk, follow-up therapy, and determine therapeutic efficacy in CVDs candidates. Based on the released guidelines, cardiac troponins (cTns) are used as a highly-sensitive and accurate approach for detecting MI. Despite the inherent advantages, the high-rate sensitivity of cTn-based assays has also led to more false positive results [@17], necessiating the advent and development of new modalities with pathological values. To improve diagnostic value upon existing MI biomarkers, the combination of complementary biological markers, such as microRNAs (miRNAs) and other genetic factors, is proposed. Previous researches support the notion that miRNAs exhibit great potential to be used as alternative biomarkers in CVDs detection and follow-up [@miR]. It is suggested that miRNAs possess 18-22 nucleotides and can play a crucial role in the regulation of gene expression. Evidence point to the fact that miRNAs are involved in the pathogenesis of cardiac tissue injury [@m1]. Several biological activities, such as angiogenesis, cardiomyocyte growth and contractility, lipid metabolism, plaque formation, and cardiac rhythm, are regulated by miRNAs [@m2]. These elements can easily circulate in biofluids and could be considered theranostics targets in terms of CVDs [@m1].  It is postulated that the function and diagnostic properties of miRNAs are beyond the myocardium in CVD patients. To be specific, the expression of miRNAs can vary in different biofluids and cell components such as serum and peripheral blood mononuclear cells (PBMCs) [@36].

PBMCs are a fraction of white blood cells, including monocytes, lymphocytes, macrophages, and other cells belonging to the immune system [@PBMC-miR]. Emerging data have indicated that PBMCs can be used as a valid source of biomarkers for monitoring various pathological conditions. Of note, the alteration of mRNAs and miRNAs under pathological conditions gives us valuable information about different kinds of disorders. PBMCs could recapitulate the conditions of the target tissues, thus, providing a highly sensitive and specific source of biomarkers [@Meder6]. Commensurate with these conditions, these cells are repositories of dysregulated genes and miRNAs expression profiles in CVDs [@PBMC-miR; @Meder6].

In recent years, the advent and application of machine learning (ML) is an exciting prospect for advancing scientific discoveries. Although the concept of ML and its initial algorithms were conceived many years ago, recent improvements in computing power and access to vast amounts of data have shown that ML techniques outperform classical statistical methods in various fields. Furthermore, the progress made in omics technologies has enabled the analysis of massive and intricate biological data sets, consisting of hundreds to thousands of samples, which makes it possible for ML to extract valuable biological insight and information from such data [@ML-Bio]. As a result, ML provides innovative methods for merging and interpreting diverse types of omics data, leading to the identification of new biomarkers. These biomarkers can aid in precise disease prediction, patient stratification, and developing novel therapeutic approaches [@ML].

In this study, we aimed to identify potential miRNA biomarkers for MI patients by combining and analyzing three different microarray datasets from PBMCs. It is suggested that the integration of omics data with bioinformatics and ML techniques could be a promising tool in the discovery of new and more accurate biomarkers for monitoring MI. Besides, this approach can deepen our vision into the underlying mechanisms of MI and aid in the development of valid diagnostic biomarkers and patient stratification.

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO) database (<https://www.ncbi.nlm.nih.gov/geo/>). To obtain sufficient classification power between MI, healthy, and CAD samples, a relatively large sample size was required. Therefore, GSE59867 for MI and CAD samples and GSE56609 and GSE54475 for healthy samples were selected. All samples were produced using Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. Only healthy, CAD, and early-stage MI samples were selected from these datasets for further analyses. The basic information for the three datasets evaluated in the current study is provided in Table \@ref(tab:datasets). The bioinformatics sections were fully conducted on R, ver. 4.2.0 [@R], using RStudio [@RStudio], and all plots and graphics of these sections were created using the ggplot2 R package [@ggplot].

@@

kbl(df, format = "markdown", booktabs = T, align = "c", linesep = c(""), caption = "Samples information on the GEO microarray datasets.")
```

## Pre-processing

The raw data in the form of CEL files from all datasets were obtained from the GEO. To prepare the data for analysis, we utilized the fRMA package [@fRMA], facilitating the pre-processing of individual microarray samples and their consistent combination. For each dataset, background correction was applied using the RMA algorithm, followed by quantile normalization based on the reference distribution. To account for probe-specific effects, batch effects were eliminated during summarization, and gene expression variances were estimated accordingly. In cases where multiple probe sets matched the same gene, the mean log fold change was retained. Consequently, fRMA can serves as a technique for removing batch effects across diverse datasets generated by identical microarray platforms [@BER]. To ensure the effectiveness of batch effect removal, we employed principal component analysis (PCA) and relative log expression (RLE) plots to visualize the data before and after applying fRMA.

@@

The barcode algorithm, as introduced by McCall et al. [@Barcode], aimed to convert actual expression values into binary barcode values. Extensive sample collections were gathered, and normalization was performed using fRMA across multiple platforms, including the Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. By utilizing these normalized datasets, the distribution of observed intensities, both for expressed and unexpressed genes, was estimated. The determination of whether a gene was expressed or unexpressed was based on the following equation, where a value of 1 indicated expression and a value of 0 indicated non-expression:

@@

In the barcode algorithm, the normalized intensity of gene $i$ in sample $j$ is denoted by $x_{ij}$. A user-defined parameter, $C$, is introduced, along with the standard deviation ($\sigma^{ne}$) and mean ($\mu^{ne}$) of the non-expressed distribution. Based on these values, the barcode representation of a sample is generated as a vector consisting of ones and zeros, representing the estimated expression (ones) and non-expression (zeros) of each gene. The barcode function within the R fRMA package was employed for implementing the barcode algorithm, utilizing the default value of $C$.

To assess the differences in expressed ratios between the MI group and the healthy control group, Fisher's exact test was performed on the barcode values of individual genes. Genes that exhibited a false discovery rate (FDR) below $0.05$, calculated using the Benjamini-Hochberg procedure to account for multiple testing issues, were identified as differentially expressed genes (DEGs). The same procedures were applied to the CAD versus healthy control comparison, as well as the MI versus CAD group, to identify DEGs specific to each comparison.

@@

The R clusterProfiler package [@cluster] was utilized to perform the Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway enrichment analysis and Gene Ontology (GO) functional annotation on the set of DEGs. The GO analysis encompassed three categories: biological process (BP), cellular component (CC), and molecular function (MF). For statistical significance, an adjusted p-value threshold of less than 0.05 was employed. The enrichment analyses were conducted separately for the DEGs specific to the MI-healthy and CAD-healthy comparisons. In these analyses, all default parameters provided by the package were used.

@@

The ML analysis was performed using Python software, ver. 3.9, Numpy [@numpy], pandas [@pandas], and Scikit-Learn packages [@scikit-learn]. Whenever hyper-tuning was needed, the scikit-opt package [@scikitopt] was used. In all ML analyses, the datasets were divided into train and test sets by a 0.7:0.3 ratio and, all reported results are the average of 10-fold cross-validation.

Two different approaches were used for selecting miRNAs for model training. The first approach was using the miRNAs that are differentially expressed. In the second approach, miRNAs with individual AUC-ROC over 0.8 for separating MI from CAD were selected. Having the result of these two different approaches can provide an informative comparison between the predictive capabilities of sets of miRNAs selected with different logics.

@@

In this approach, a two-layer architecture was deployed to the data to maximize the prediction values. The first layer predicted whether a sample was healthy or not, and the second layer separated MI from CAD in the samples which were predicted as not healthy in the first layer. To this end, a distinct ML model was trained for each layer. Since there is a limited number of miRNAs in DEGs, both layers were trained with all of them. For further comparison with the models' performance, the ROC curve of each miRNA for classifying healthy and not-healthy, as well as CAD and MI, were generated using a Logistic Regression model.

#### First layer for isolation of healthy and not-healthy samples:

A support vector machine (SVM) model using RBF kernels was trained and hyper-tuned using all miRNAs in DEGs. To handle the severe imbalance in the number of samples (51 for the healthy and 157 for the not-healthy groups), the sample weight for the healthy and the not-healthy samples was set to 1 and 0.5, respectively. The ROC curve and confusion matrix for the model were reported.

#### Second layer for separating MI and CAD samples:

For the sake of reaching the highest classification performance, different models were investigated. To do so, SVM (with linear, polynomial, and RBF kernels), Logistic Regression (LR), Random Forests (RF), k-Nearest Neighbor (kNN), Gradient Boosting (GB), XGBoost (XGB) and Decision Tree (DT) models were trained. All models were trained with their pre-set parameters with 10-fold cross-validation. The criteria for choosing the best model were the highest accuracy and AUC-ROC on the test set. The best model was hyper-tuned with the scikit-opt package [@scikitopt] for the best classification performance. The ROC curve and confusion matrix for the best model were reported.

@@

Like the previous approach, a two-layers strategy was conducted. The first layer classified samples into healthy and not-healthy, and the separated MI and CAD samples. However, to keep the number of miRNAs as low as possible, miRNAs were selected from the second layer, and then their performance was evaluated in the first layer. The AUC-ROC of all miRNAs for classifying MI and CAD samples were calculated and, the miRNAs with the AUC-ROC over 0.8 were selected. The ROC curves for each selected miRNA for separating healthy samples from not-healthy and MI from CAD samples were also plotted for further comparison.

#### First layer for isolation of healthy and not-healthy samples:

An SVM model with an RBF kernel was trained using the selected set of miRNAs. Additionally, the model was hyper-tuned to find the hyper-parameters for the highest AUC-ROC and accuracy. The same sample weights as the previous approach (1 for healthy and 0.5 for not-healthy samples) were used. The ROC curve and confusion matrix for the model were reported.

#### Second layer for separating MI and CAD samples:

The selected miRNA set was used to train different algorithms to find the best model. Similar to the previous approach, SVM (with linear, polynomial, and RBF kernels), LR, RF, kNN, GB, XGB, and DT models were trained. All models were trained with their pre-set parameters using 10-fold cross-validation. The models with the highest AUC-ROC and accuracy on the test set were selected and hyper-tuned using the scikit-opt package [@scikitopt]. The ROC curve and confusion matrix for the best model were reported.

@@

## Pre-processing

The PCA plots of the samples are shown in Figures \@ref(fig:PCA)A and B. As shown, healthy samples were separated from CAD or MI samples in primary data and after conducting fRMA. In the RLE plot, there was a distinct difference between dataset means for all samples before performing fRMA (Figure \@ref(fig:PCA)C). All datasets were rearranged around 0 in the RLE plot after conducting fRMA (Figure \@ref(fig:PCA)D). Moreover, there was an apparent change in inter-quantile distances, but the values still were over 0.1.

```{r PCA, echo=F, out.width="100%", fig.cap="principal component analysis plots for (A) primary data and (B) the data after fRMA, and the relative log expression plots for (C) primary data and (D) the data after fRMA."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/PCA and RLE of trains.pdf")
```

@@

According to the cutoff criterion of $FDR < 0.05$, there were `r dim(M_H)[1]` DEGs between the MI and the healthy samples. Among them, `r sum(M_H$logFC > 0)` were up-regulated, and `r sum(M_H$logFC < 0)` were down-regulated in MI compared to the healthy controls. In CAD and healthy group comparison, we found `r dim(C_H)[1]` DEGs, of which `r sum(C_H$logFC > 0)` and `r sum(C_H$logFC < 0)` DEGs were up- and down-regulated, respectively, in CAD samples. In the MI and CAD groups, the number of DEGs was `r dim(M_C)[1]`, and the number of up- and down-regulated genes in MI samples were `r sum(M_C$logFC > 0)` and `r sum(M_C$logFC < 0)` compared to CAD samples, respectively. These data are summarized in Table \@ref(tab:DEGstab).

```{r venn, echo=F, out.width="45%", fig.align='center', fig.cap="Venn diagram for DEGs in CAD/Healthy, MI/Healthy, and MI/CAD comparison."}

@@

The Venn diagram in Figure \@ref(fig:venn) shows that CAD and MI samples shared most of their DEGs. From 860 DEGs of MI/healthy and 670 DEGs of CAD/healthy, 531 genes were common, which is 62% of MI/healthy DEGs and 79% of CAD/healthy DEGs.

@@

To explore the biological classification of the DEGs, we performed GO and KEGG pathway enrichment analyses on MI/healthy and CAD/healthy DEGs. For MI/healthy, GO enrichment analysis in the BP category suggested that the DEGs were enriched in "immune response-regulating signaling pathway," "lymphocyte differentiation," "immune response-regulating cell surface receptor signaling pathway," and "leukocyte activation involved in immune response" (Figure \@ref(fig:MIHEnrich)A). In the CC category, the DEGs were enriched in "secretory granule membrane," "azurophil granule," "ficolin-1-rich granule," "tertiary granule," and "ficolin-1-rich granule membrane" (Figure \@ref(fig:MIHEnrich)B). In the MF category, the DEGs were involved in "cadherin binding" and "MHC class I protein binding" (Figure \@ref(fig:MIHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway," "Lipid and atherosclerosis," and "Hematopoietic cell lineage" (Figure \@ref(fig:MIHEnrich)D).

@@

The enrichment results for CAD/healthy DEGs were as follows. In the BP category, GO enrichment suggested that the DEGs were enriched in "positive regulation of defense response," "positive regulation of innate immune response," "mononuclear cell differentiation," and "positive regulation of response to external stimulus" (Figure \@ref(fig:CADHEnrich)A). In the CC category, the DEGs were enriched in "azurophil granule," "ficolin-1-rich granule," and "ficolin-1-rich granule membrane" (Figure \@ref(fig:CADHEnrich)B). In the MF category, the DEGs were involved in "lipoprotein particle receptor binding" and "NF-$\kappa$B binding" (Figure \@ref(fig:CADHEnrich)C). KEGG pathway analysis showed that the DEGs were related to the following pathways: "Chemokine signaling pathway," "Lipid and atherosclerosis," and "Hematopoietic cell lineage" (Figure \@ref(fig:CADHEnrich)D).

@@

Among all DEGs, just miR-186, miR-32, and miR-21 were detected as differentially expressed miRNAs. The expression profile of these three miRNAs is presented in Figure \@ref(fig:Expall). Additionally, The ROC curves of each miRNA for each layer are presented in Figure \@ref(fig:miRROC). Using the logistic regression model, the AUC-ROC of miR-21, miR-32, and miR-186 for separating healthy and not-healthy samples was 0.98, 0.99, and 0.90, respectively (Figure \@ref(fig:miRROC)A). Besides, the accuracy of each miRNA for classifying the samples into healthy and not-healthy groups on the test set was 0.92, 0.98, and 0.89 for miR-21, miR-32, and miR-186, respectively. Moreover, the ROC curve of each miRNA for classifying MI and CAD samples was presented in Figure \@ref(fig:miRROC)B. The AUC-ROC and accuracy for miR-21, miR-32, and miR-186 on the test set were 0.85; 0.70; and 0.86, and 0.78; 0.67; and 0.74, respectively.

@@

#### First layer for healthy not-healthy isolation:

>>>>

---
title: Machine Learning Driven Set of microRNAs as a Novel Biomarker Set for Myocardial Infarction Diagnosis
authors:
  - name: "Mehrdad Samadishadlou"
    address: "Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran"
date: "`r Sys.Date()`"
link-citation: yes
bibliography: [For Article Refrence.bib, discussion.bib, Introduction.bib, mir296.bib]
abstract: |
  MicroRNAs (miRNAs) play a crucial role in regulating adaptive and maladaptive responses in cardiovascular diseases, making them attractive targets for potential biomarkers. However, their potential as novel biomarkers for diagnosing cardiovascular diseases requires systematic evaluation. In this study, we aimed to identify a key set of miRNA biomarkers using integrated bioinformatics and machine learning analysis. We combined and analyzed three gene expression datasets from the Gene Expression Omnibus (GEO) database, which contains peripheral blood mononuclear cell (PBMC) samples from individuals with myocardial infarction (MI), stable coronary artery disease (CAD), and healthy individuals. Additionally, we selected a set of miRNAs based on their area under the receiver operating characteristic curve (AUC-ROC) for separating the CAD and MI samples. We designed a two-layer architecture for sample classification, in which the first layer isolates healthy samples from unhealthy samples, and the second layer classifies stable CAD and MI samples. We trained different machine learning models using both biomarker sets and evaluated their performance on a test set. We identified miR-21, miR-186, and miR-32 as the only miRNAs among the differentially expressed genes, and a set including miR-186, miR-21, miR-197, miR-29A, and miR-296 as the optimum set of miRNAs selected by their AUC-ROC. Both biomarker sets could distinguish healthy from not-healthy samples with complete accuracy. The best performance for the classification of CAD and MI was achieved with an SVM model trained using the biomarker set selected by AUC-ROC, with an AUC-ROC of 0.96 and an accuracy of 0.94 on the test data. Our study demonstrated that miRNA signatures derived from PBMCs could serve as valuable novel biomarkers for cardiovascular diseases.

output: 
  bookdown::word_document2: default
  bookdown::pdf_book:
always_allow_html: true
---

@@

Cardiovascular diseases (CVDs) are the leading cause of human mortality, accounting for 32% of all global deaths. It is estimated that approximately 85% of CVD mortality is due to myocardial infarction (MI) [@CVD]. MI is an acute coronary syndrome characterized by sudden blockage and stenosis of the coronary artery and subsequent myocardial ischemia, leading to extensive cardiomyocyte damage and necrosis [@Macro].

Over the last 50 years, numerous attempts have been made to use biomarkers to facilitate diagnosis, assess the risk, follow-up therapy, and determine therapeutic efficacy in CVD candidates. Based on released guidelines, cardiac troponins (cTns) are used as a highly sensitive and accurate approach for detecting MI. Despite these inherent advantages, the high sensitivity of cTn-based assays has also led to more false-positive results [@17], necessitating the advent and development of new modalities with pathological value. To improve the diagnostic value of  existing MI biomarkers, a combination of complementary biological markers, such as microRNAs (miRNAs) and other genetic factors, has been proposed. Previous research supports the notion that miRNAs exhibit great potential as alternative biomarkers for CVD detection and follow-up [@miR]. It has been suggested that miRNAs possess 18-22 nucleotides and play a crucial role in the regulation of gene expression. Evidence indicates that miRNAs are involved in the pathogenesis of cardiac tissue injury [@m1]. Several biological processes, such as angiogenesis, cardiomyocyte growth and contractility, lipid metabolism, plaque formation, and cardiac rhythm, are regulated by miRNAs [@m2]. These elements can easily circulate in biofluids and could be considered theranostic targets in terms of CVDs [@m1].  It has been postulated that the function and diagnostic properties of miRNAs are beyond the myocardium in patients with CVD.  Specifically, the expression of miRNAs can vary in different biofluids and cell components such as serum and peripheral blood mononuclear cells (PBMCs) [@36].

PBMCs are a fraction of white blood cells, including monocytes, lymphocytes, macrophages, and other cells of the immune system [@PBMC-miR]. Emerging data indicate that PBMCs can be used as a valid source of biomarkers for monitoring various pathological conditions. Of note, the alteration of mRNAs and miRNAs under pathological conditions provides valuable information about different kinds of disorders. PBMCs can recapitulate the conditions of target tissues, thus providing a highly sensitive and specific source of biomarkers [@Meder6]. Combined with these conditions, these cells are repositories of dysregulated genes and miRNA expression profiles in CVDs [@PBMC-miR; @Meder6].

In recent years, the advent and application of machine learning (ML) has been an exciting prospect for advancing scientific research. Although the concept of ML and its initial algorithms were conceived many years ago, recent improvements in computing power and access to vast amounts of data have demonstrated that ML techniques outperform classical statistical methods in various fields. Furthermore, the progress made in omics technologies has enabled the analysis of massive and intricate biological datasets, consisting of hundreds to thousands of samples, which makes it possible for ML to extract valuable biological insights and information from such data [@ML-Bio]. Consequently, ML provides innovative methods for merging and interpreting diverse types of omics data, leading to the identification of new biomarkers. These biomarkers can aid in precise disease prediction, patient stratification, and the development of novel therapeutic approaches [@ML].

In this study, we aimed to identify potential miRNA biomarkers in patients with MI by combining and analyzing three different microarray datasets from PBMCs. The integration of omics data with bioinformatics and ML techniques could be a promising tool in the discovery of new and more accurate biomarkers for monitoring MI. Additionally, this approach can deepen our understanding of the underlying mechanisms of MI and aid in the development of valid diagnostic biomarkers and patient stratification.

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO) database (<https://www.ncbi.nlm.nih.gov/geo/>). To obtain sufficient classification power between MI, healthy, and CAD samples, a relatively large sample size is needed Therefore, GSE59867 for the MI and CAD samples and GSE56609 and GSE54475 for the healthy samples were selected. All samples were produced using Affymetrix Human Gene 1.0 ST Array platform (GPL6244). Only healthy, CAD, and early-stage MI samples were selected from these datasets for further analysis. The basic information for the three datasets evaluated in this study is provided in Table \@ref(tab:datasets). Bioinformatics analyses were conducted using R, ver. 4.2.0 [@R], RStudio [@RStudio]. All plots and graphics of these sections were created using the ggplot2 R package [@ggplot].

@@

kbl(df, format = "markdown", booktabs = T, align = "c", linesep = c(""), caption = "Sample information on the GEO microarray datasets.")
```

## Preprocessing

The raw data in the form of CEL files from all datasets were obtained from GEO. To prepare the data for analysis, we utilized the fRMA package [@fRMA] to facilitate preprocessing of individual microarray samples and their consistent combination. For each dataset, background correction was applied using the RMA algorithm, followed by quantile normalization based on the reference distribution. To account for probe-specific effects, batch effects were eliminated during summarization and gene expression variances were estimated accordingly. In cases where multiple probe sets matched the same gene, the mean log-fold change was retained. Consequently, fRMA can serve as a technique to remove batch effects across diverse datasets generated by identical microarray platforms [@BER]. To ensure the effectiveness of the batch effect removal, we employed principal component analysis (PCA) and relative log expression (RLE) plots to visualize the data before and after applying fRMA.

@@

The barcode algorithm was introduced by McCall et al. [@Barcode], aimed to convert actual expression values into binary barcode values. Extensive sample collections were gathered and normalization was performed using fRMA across multiple platforms, including the Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. By utilizing these normalized datasets, the distribution of the observed intensities for both the expressed and unexpressed genes was estimated. The determination of whether a gene was expressed or not was based on the following equation, where a value of 1 indicates expression and a value of 0 indicates non-expression:

@@

In the barcode algorithm, the normalized intensity of gene $i$ in sample $j$ is denoted as $x_{ij}$. A user-defined parameter, $C$, was introduced along with the standard deviation ($\sigma^{ne}$) and mean ($\mu^{ne}$) of the non-expressed distribution. Based on these values, the barcode representation of a sample was generated as a vector consisting of ones and zeros, representing the estimated expression (ones) and non-expression (zeros) of each gene. The barcode function within the R fRMA package was employed to implement the barcode algorithm, utilizing the default value of $C$.

To assess the differences in expressed ratios between the MI and healthy control groups, Fisher's exact test was performed on the barcode values of individual genes. Genes that exhibited a false discovery rate (FDR) below $0.05$, calculated using the Benjamini-Hochberg procedure to account for multiple testing issues were identified as differentially expressed genes (DEGs). The same procedures were applied to the CAD versus healthy control comparison, as well as to the MI versus CAD group, to identify DEGs specific to each comparison.

@@

The R clusterProfiler package [@cluster] was utilized to perform Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway enrichment analysis and Gene Ontology (GO) functional annotation on the set of DEGs. GO analysis encompassed three categories: biological process (BP), cellular component (CC), and molecular function (MF). For statistical significance, an adjusted p-value threshold of less than 0.05 was employed. Enrichment analyses were conducted separately for DEGs specific to the MI-healthy and CAD-healthy comparisons. All the default parameters provided by the package were used in the analyses.

@@

ML analysis was performed using Python software, ver. 3.9, Numpy [@numpy], Pandas [@pandas], and Scikit-Learn packages [@scikit-learn]. Whenever hypertuning was needed, the Scikit-opt package [@scikitopt] was used. In all ML analyses, the datasets were divided into training and test sets at a 0.7:0.3 ratio, and all reported results are the average of 10-fold cross-validation.

Two different approaches were used to select miRNAs for model training. The first approach was to use differentially expressed miRNAs. In the second approach, miRNAs with individual AUC-ROCs over 0.8 for separating MI from CAD were selected. The results of these two approaches can provide an informative comparison between the predictive capabilities of sets of miRNAs selected with different logics.

@@

In this approach, a two-layer architecture is deployed to the data to maximize the prediction values. The first layer predicted whether a sample was healthy or not, and the second layer separated MI from CAD in the samples that were predicted as not healthy in the first layer. To this end, a distinct ML model was trained for each layer. Because, there were a limited number of miRNAs in the DEGs, both layers were trained with all of them. For further comparison with the models' performance, the ROC curve of each miRNA for classifying healthy and not-healthy, as well as CAD and MI, was generated using a logistic regression model.

#### First layer for the isolation of healthy and not-healthy samples:

A support vector machine (SVM) model using RBF kernels was trained and hypertuned using all miRNAs in the DEGs. To handle the severe imbalance in the number of samples (51 for the healthy group and 157 for the not-healthy group), the sample weights for the healthy and the not-healthy samples were set to 1 and 0.5, respectively. The ROC curve and confusion matrix for the model are reported.

#### Second layer for separating the MI and CAD samples:

Different models were investigated to achieve the highest classification performance. To do so, SVM (with linear, polynomial, and RBF kernels), logistic regression (LR), random forests (RF), k-nearest neighbor (kNN), gradient boosting (GB), XGBoost (XGB) and decision tree (DT) models were trained. All models were trained with their preset parameters using 10-fold cross-validation. The criteria for selecting the best model were the highest accuracy and AUC-ROC for the test set. The best model was hypertuned using the scikit-opt package [@scikitopt] for the best classification performance. The ROC curve and confusion matrix for the best model are reported.

@@

As in the previous approach, a two-layer strategy was employed. The first layer classified samples into healthy and not-healthy, and the second layer separated the MI and CAD samples. However, to keep the number of miRNAs as low as possible, miRNAs were selected from the second layer and their performance was evaluated in the first layer. The AUC-ROC of all miRNAs for classifying MI and CAD samples was calculated, and miRNAs with AUC-ROC $> 0.8$ were selected. ROC curves for each selected miRNA for separating healthy samples from not-healthy samples and MI from CAD samples were also plotted for further comparison.

#### First layer for the isolation of healthy and not-healthy samples:

An SVM model with an RBF kernel is trained using the selected set of miRNAs. Additionally, the model was hypertuned to find the hyperparameters for the highest AUC-ROC and accuracy. The same sample weights as in the previous approach (1 for healthy and 0.5 for not-healthy samples) were used. The ROC curve and confusion matrix for the model were reported.

#### Second layer for separating the MI and CAD samples:

The selected miRNA set was used to train different algorithms to determine the best model. Similar to the previous approach, the SVM (with linear, polynomial, and RBF kernels), LR, RF, kNN, GB, XGB, and DT models were trained. All models were trained with their preset parameters using 10-fold cross-validation. The models with the highest AUC-ROC and accuracy on the test set were selected and hypertuned using the scikit-opt package [@scikitopt]. The ROC curve and confusion matrix for the best model were reported.

@@

## Preprocessing

The PCA plots of the samples are shown in Figures \@ref(fig:PCA)A and B. Healthy samples were separated from the CAD or MI samples in the primary data and after conducting fRMA. In the RLE plot, there was a distinct difference between the dataset means for all samples before fRMA was performed (Figure \@ref(fig:PCA)C). All datasets were rearranged to approximately 0 in the RLE plot after fRMA was conducted (Figure \@ref(fig:PCA)D). Moreover, there was an apparent change in the interquantile distances, but the values were still greater than 0.1.

```{r PCA, echo=F, out.width="100%", fig.cap="Principal component analysis plots for (A) primary data and (B) the data after fRMA, and the relative log expression plots for (C) primary data and (D) the data after fRMA."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/PCA and RLE of trains.pdf")
```

@@

According to the cutoff criterion of $FDR < 0.05$, there were `r dim(M_H)[1]` DEGs between MI and healthy samples. Among them, `r sum(M_H$logFC > 0)` were up-regulated, and `r sum(M_H$logFC < 0)` were down-regulated in the MI group compared to the healthy group. In the CAD and healthy group comparison, we found `r dim(C_H)[1]` DEGs, of which `r sum(C_H$logFC > 0)` and `r sum(C_H$logFC < 0)` DEGs were up- and down-regulated, respectively, in CAD samples. In the MI and CAD groups, the number of DEGs was `r dim(M_C)[1]`, and the numbers of up- and down-regulated genes in MI samples were `r sum(M_C$logFC > 0)` and `r sum(M_C$logFC < 0)`, respectively, compared to CAD samples. The data are summarized in Table \@ref(tab:DEGstab).

```{r venn, echo=F, out.width="45%", fig.align='center', fig.cap="Venn diagram for DEGs in CAD/Healthy, MI/Healthy, and MI/CAD."}

@@

The Venn diagram in Figure \@ref(fig:venn) shows that the CAD and MI samples shared most of their DEGs. From 860 DEGs of MI/healthy and 670 DEGs of CAD/healthy, 531 genes were common, which is 62% of MI/healthy DEGs and 79% of CAD/healthy DEGs.

@@

To explore the biological classification of the DEGs, we performed GO and KEGG pathway enrichment analyses on the MI/healthy and CAD/healthy DEGs. For MI/healthy, GO enrichment analysis in the BP category suggested that the DEGs were enriched in "immune response-regulating signaling pathway," "lymphocyte differentiation," "immune response-regulating cell surface receptor signaling pathway," and "leukocyte activation involved in immune response" (Figure \@ref(fig:MIHEnrich)A). In the CC category, DEGs were enriched in "secretory granule membrane," "azurophil granule," "ficolin-1-rich granule," "tertiary granule," and "ficolin-1-rich granule membrane" (Figure \@ref(fig:MIHEnrich)B). In the MF category, DEGs were involved in "cadherin binding" and "MHC class I protein binding" (Figure \@ref(fig:MIHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway," "Lipid and atherosclerosis," and "Hematopoietic cell lineage" (Figure \@ref(fig:MIHEnrich)D).

@@

The enrichment results for the CAD/healthy DEGs were as follows. In the BP category, GO enrichment suggested that the DEGs were enriched in "positive regulation of defense response," "positive regulation of innate immune response," "mononuclear cell differentiation," and "positive regulation of response to external stimulus" (Figure \@ref(fig:CADHEnrich)A). In the CC category, DEGs were enriched in "azurophil granule," "ficolin-1-rich granule," and "ficolin-1-rich granule membrane" (Figure \@ref(fig:CADHEnrich)B). In the MF category, DEGs were involved in "lipoprotein particle receptor binding" and "NF-$\kappa$B binding" (Figure \@ref(fig:CADHEnrich)C). KEGG pathway analysis showed that the DEGs were related to the following pathways: "Chemokine signaling pathway," "Lipid and atherosclerosis," and "Hematopoietic cell lineage" (Figure \@ref(fig:CADHEnrich)D).

@@

Among the DEGs, miR-186, miR-32, and miR-21 were identified as differentially expressed miRNAs. The expression profiles of the three miRNAs are shown in Figure \@ref(fig:Expall). The ROC curves of each miRNA in each layer are presented in Figure \@ref(fig:miRROC). Using the logistic regression model, the AUC-ROC values of miR-21, miR-32, and miR-186 for separating healthy and not-healthy samples were 0.98, 0.99, and 0.90, respectively (Figure \@ref(fig:miRROC)A). The accuracy of each miRNA for classifying the samples into healthy and not-healthy groups on the test set for miR-21, miR-32, and miR-186 was 0.92, 0.98, and 0.89, respectively. The ROC curve of each miRNA for classifying MI and CAD samples is presented in Figure \@ref(fig:miRROC)B. The AUC-ROC and accuracy for miR-21, miR-32, and miR-186 in the test set were 0.85; 0.70; and 0.86, and 0.78; 0.67; and 0.74, respectively.

@@

#### First layer for the isolation of healthy and not-healthy samples:

Although single miRNAs had an acceptable performance for this layer, their predictive value could be  further improved by using them as a set. The ROC curve for the SVM model with an RBF kernel trained with all three miRNAs is presented in Figure \@ref(fig:DEMROC)A. The model had a better performance in classification than single miRNAs.The AUC-ROC for the model was 1, and its accuracy on the test set was also 1. In Figure \@ref(fig:DEMCF)A, the confusion matrix for the model is presented.
